---
title: "Descriptive statistics"
output: 
  ioslides_presentation: 
    smaller: yes
    fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(tibble.print_min = 5, tibble.print_max = 5)
```


# Summarising variables

## Summarising a numerical variable {.smaller}

```{r, warning=FALSE, echo=TRUE, results='hide', message=FALSE}
library(nycflights13)
data(package = 'nycflights13')
```

* How is the distribution of `dep_delay`?

```{r, warning=FALSE, echo=TRUE, results='hide', message=FALSE}
summarise_dep_delay = function(.data) 
  summarise(.data,
            mean_ = mean(dep_delay, na.rm = TRUE),
            median_ = median(dep_delay, na.rm = TRUE),
            q1_ = quantile(dep_delay, 0.25, na.rm = TRUE),
            q3_ = quantile(dep_delay, 0.75, na.rm = TRUE),
            iqr_ = IQR(dep_delay, na.rm = TRUE),
            min_ = min(dep_delay, na.rm = TRUE),
            max_ = max(dep_delay, na.rm = TRUE) )
summarise_dep_delay(flights)
```

* Highly right-skewed distribution (_mean_ much higher than _median_)
* Distribution centered around zero (_median_ equal to -2)
* 50% of central observations between -5 and 11 ($Q_1$ and $Q_3$)

## Summarising by groups {.smaller}

* How is the distribution of `dep_delay` with respect to `origin`?

```{r}
flights_grouped = group_by(flights, origin)
summarise_dep_delay(flights_grouped)
```

* Distributions of `dep_delay` are similar with respect `origin`.
* Lower variability of `dep_delay` in `LGA` airport.

## Relation between categorical variables {.smaller}

Let's create an `arrival` variable with two values: "delayed" and "on-time".

```{r}
flights = mutate(flights, arrival = if_else(arr_delay > 0, 'delayed', 'on-time'))
```

* How is the distribution of `origin` and `on.time`

```{r}
tab = with(flights, table(arrival, origin))
# tab = table(flights[["arrival"]], flights[["origin"]])
# tab = table(flights$arrival, flights$origin)
```

* Show cross-tabulation with margins:

```{r}
addmargins(tab)
```

## Relation between categorical variables.  {.smaller}

__The marginals__

* Row relative frequencies

```{r}
prop.table(tab, margin = 1)
```

* Column relative frequencies

```{r}
prop.table(tab, margin = 2)
```

## Relation between numerical variables {.smaller}

* How is the distribution of `dep_delay` and `arr_delay`?

```{r}
summarise(flights,
          m_dep_delay = median(dep_delay, na.rm = TRUE),
          iqr_dep_delay = IQR(dep_delay, na.rm = TRUE),
          m_arr_delay = median(arr_delay, na.rm = TRUE),
          iqr_arr_delay = IQR(arr_delay, na.rm = TRUE),
          s_dep_arr_delay = cor(dep_delay, arr_delay, use = 'complete.obs')
)
```

* Higher variability in the arrival delay.
* Strong association between `dep_delay` and `arr_delay`.

## `magrittr`: A Forward-Pipe Operator

* Suppose we are interested in selecting columns: `month`, `day`, `arr_delay`, `dep_delay`; keeping only observations from 02/02/2013.

    * Using `select()`and `filter()` we would proceed as follows:
    
    ```{r, results = 'hide'}
    tab = mutate(flights, month, day, arr_delay, dep_delay)
    filter(tab, month == 2, day == 13)
    ```

    * The pipe operator, `%>%`,  simplifies the notation and avoids creating temporal tables.
    
    ```{r, results = 'hide'}
    flights %>%
      mutate(month, day, arr_delay, dep_delay) %>%
      filter(month == 2, day == 13)
    ```

## Dataset importation {.smaller data-background=#CCE5FF}

* Import the dataset `birthwt: Risk Factors Associated with Low Infant Birth Weight` available at `birthwt.xlsx` in `RSTAT-INTRO-SESSION-3` project. 

  This dataset contains the following columns:
  
    * `low` indicator of birth weight less than 2.5 kg.
    * `age` mother's age in years.
    * `lwt` mother's weight in pounds at last menstrual period.
    * `race` mother's race (1 = white, 2 = black, 3 = other).
    * `smoke` smoking status during pregnancy.
    * `ptl` number of previous premature labours.
    * `ht` history of hypertension.
    * `ui` presence of uterine irritability.
    * `ftv` number of physician visits during the first trimester.
    * `bwt` birth weight in grams.

## Summarising variables {.smaller data-background=#CCE5FF}

* Express mother's weight in Kg.
* Summarise infant and mother's weight.
* Get the frequencies between smoke and low indicator variables. Draw some conclusion from obtained frequencies.
* Calculate the average of infant's weight by smoking, race and hypertension mother's conditions.


<!-- # Wrangling

## Tidy Data

Tidy data is a standard way of mapping the meaning of a dataset to its structure.

1. Each variable forms a column
2. Each observation forms a row
3. Each type of observational unit forms a table

## `tidyr`: helpers to create tidy data

* "Pivotting". `pivot_longer()` and `pivot_wider()`
* "Rectangling".
* "Nesting".
* Splitting and combining: `separate()` and `unite()`.

<br><br><br><br>

Cheat Sheet [pdf](https://github.com/rstudio/cheatsheets/raw/master/data-import.pdf)

## Messy datasets (1) {.smaller} 

__Column headers are values, no variable names__

```{r}
tidyr::relig_income
```

## Messy datasets (2) {.smaller}   

__Multiple variables stored in a column__

```{r}
tidyr::who
```

## Messy datasets (3) {.smaller} 

__Variables are stored in both rows and columns__

```{r, message=FALSE, eval=FALSE}
library(readr)
read_csv("session-02-presentation/weather.csv")
```

<!-- ## About correlation

* Correlation can be easily misinterpreted.
  * To measure the effect, we need to be free of confounding. [Causal inference course](https://www.bradyneal.com/causal-inference-course)
  * Correlation is not valid for relative data.
      * Variation array for compositional data.
      * Proportionality: A Valid Alternative to Correlation for Relative Data. [Reference](https://pubmed.ncbi.nlm.nih.gov/25775355/)

-->